Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-14 main.tex
Unification nets (part 1)

\newcommand{\Gr}{\mathcal{G}}
\newcommand{\FT}{\mathcal{F}}
\newcommand{\US}{\mathcal{U}}
\newcommand{\PE}[1]{#1^P}

@@

\newcommand{\form}[1]{\mathrm{fm}(#1)}

@@

\begin{figure}[!t]
  \begin{center}
    \framebox{%
      $
      \begin{array}{@{}c@{}}
        \dbox{%
          $
          \begin{array}{c}
            \vlinf{\axr}{}{\sqn{a,\cneg a}}{}
            \qquad
            \vlinf{\ttt}{}{\sqn{\ttt}}{}
            \qquad
            \vliinf{\mixr}{}{\sqn{\Gamma,\Delta}}{\sqn\Gamma}{\sqn\Delta}
            \\ \\[-1ex]
            \vlinf{\vlor}{}{\sqn{\Gamma,A\vlor B}}{\sqn{\Gamma,A,B}}
            \qquad
            \vliinf{\vlan}{}{\sqn{\Gamma,A\vlan B}}{\sqn{\Gamma,A}}{\sqn{B,\Delta}}
            \\ \\[-1ex]
            \vlinf{\exists}{}{\sqn{\Gamma,\exists x.A}}{\sqn{\Gamma,A\ssubst{x}{t}}}
            \qquad
            \vlinf{\forall}{\text{($x$ not free in $\Gamma$)}}{\sqn{\Gamma,\forall x.A}}{\sqn{\Gamma,A}}
          \end{array}
          $%
        }%
        \\\\[-1ex]
        \vlinf{\conr}{}{\sqn{\Gamma,A}}{\sqn{\Gamma,A,A}}
        \qquad
        \vlinf{\weakr}{}{\sqn{\Gamma,A}}{\sqn{\Gamma}}
      \end{array}
      $
    }
  \end{center}
  \caption{Sequent calculi $\FOLK$ (all rules) and $\FOMLL$ (rules in the dashed box)}
  \label{fig:LK1}
\end{figure}

@@

\todo{}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Translation between Unification Nets and $\FOMLL$}

\todo{}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Translation between Unification Nets and Fonets}

@@

In $\FOMLL$, terms, atoms, formulas are defined as in first-order logic. For
simplicity, we choose to use $\cor$ and $\cand$ instead of $\parr$ and
$\otimes$ which are generally used in the presentation of linear logic.
A formula $A$ is identified with its \bfit{formula tree} $\FT(A)$, a directed
tree with leaves labelled by atoms, internal nodes labelled by connectives and
quantifiers, and edges directed from leaves to the root. A \bfit{sequent} $\Gamma$ is simply a disjoint union of formulas. We write
comma for disjoint union.

@@

Let $\lambda$ is a unifiable linking on $\Gamma = A_1, \ldots, A_n$.
Let $A = A_1 \cor \ldots \cor A_n$. The \bfit{unification structure}
$\US(\lambda)$
associated to $\lambda$ is the formula tree $\FT(A)$ together with an undirected
edge between leaves $l$ and $l'$ for every link $\{l, l'\}$ in $\lambda$ and a
directed edge from $\exists x$ to $\forall y$ for every dependency $\exists x
\rightarrow \forall y$.

A \bfit{switching graph} of a unification structure $\US(\lambda)$ is any
derivative of $\US(\lambda)$ obtained by keeping only one edge into each $\cor$
and $\forall$ and undirecting remaining edges. A linking is \bfit{correct} if it
is unifiable and all of the switching graphs of its associated unification
structure are acyclic.

@@

\begin{definition} A \bfit{switching path} of a  formula tree $\FT(A)$ is a
	path in $\FT(A)$ that does not go through both incoming edges of a $\cor$.
\end{definition}

@@

\begin{proposition}
\label{prop2}
Let $u$ and $v$ be two distinct vertices of a fograph $\graphof(A)$, then we have the equivalence
between:
\begin{itemize}
  \item $u$ and $v$ are adjacent in $\graphof(A)$
  \item $u$ and $v$ are connected by a switching path of $\FT(A)$, and if one of them is a universal quantifier, then the other is
	not a descendant of the former.
  \end{itemize}
\begin{proof}
By induction on $A$.
\begin{itemize}
  \item If $A$ is an atom, trivial.
  \item If $A = A_1 \cand A_2$, then we distinguish two cases:
    \begin{itemize}
      \item $u$ and $v$ are both in $A_1$ (resp. $A_2$): trivial by
	      the induction hypothesis.
      \item one of them is in $A_1$ and the other is in $A_2$: they are
	      adjacent in $\graphof(A)$ by definition. By Proposition \ref{prop1},
		    the one in $A_1$ (resp. $A_2$) is connected to the
		    vertex representing $A_1$ (resp. $A_2$) by a switching
		    path. Together with the two edges incident to $A_1 \cand
		    A_2$, we obtain a switching path connecting $u$ and $v$.
    \end{itemize}
  \item If $A = A_1 \cor A_2$, then we distinguish two cases:
    \begin{itemize}
      \item $u$ and $v$ are both in $A_1$ (resp. $A_2$): trivial by
	      the induction hypothesis.
      \item one of them is in $A_1$ and the other is in $A_2$: they are not
	      adjacent in $\graphof(A)$ by definition. It is clear that they
		    are not connected by a switching path.
    \end{itemize}
  \item If $A = \exists x \ A'$, then we distinguish two cases:
    \begin{itemize}
      \item $u$ and $v$ are both in $A'$: trivial by the induction
	      hypothesis.
      \item one of them is $\exists x$ and the other is in $A'$: trivial
	      by Proposition \ref{prop1}

>>>>

\newcommand{\Gr}{\mathcal{G}}
\newcommand{\PE}[1]{#1^P}

@@

\newcommand{\form}[1]{\mathrm{fm}(#1)}
\newcommand{\formtree}[1]{\mathcal{F}(#1)}

\newcommand{\unifstr}[1]{\mathcal{U}(#1)}

@@

\begin{figure}[!t]
  \begin{center}
    \framebox{%
      $
      \begin{array}{@{}c@{}}
        \dbox{%
          $
          \begin{array}{c}
            \vlinf{\axr}{}{\sqn{a,\cneg a}}{}
            \qquad
            \vlinf{\ttt}{}{\sqn{\ttt}}{}
            \qquad
            \vliinf{\mixr}{}{\sqn{\Gamma,\Delta}}{\sqn\Gamma}{\sqn\Delta}
            \\ \\[-1ex]
            \vlinf{\vlor}{}{\sqn{\Gamma,A\vlor B}}{\sqn{\Gamma,A,B}}
            \qquad
            \vliinf{\vlan}{}{\sqn{\Gamma,A\vlan B, \Delta}}{\sqn{\Gamma,A}}{\sqn{B,\Delta}}
            \\ \\[-1ex]
            \vlinf{\exists}{}{\sqn{\Gamma,\exists x.A}}{\sqn{\Gamma,A\ssubst{x}{t}}}
            \qquad
            \vlinf{\forall}{\text{($x$ not free in $\Gamma$)}}{\sqn{\Gamma,\forall x.A}}{\sqn{\Gamma,A}}
          \end{array}
          $%
        }%
        \\\\[-1ex]
        \vlinf{\conr}{}{\sqn{\Gamma,A}}{\sqn{\Gamma,A,A}}
        \qquad
        \vlinf{\weakr}{}{\sqn{\Gamma,A}}{\sqn{\Gamma}}
      \end{array}
      $
    }
  \end{center}
  \caption{Sequent calculi $\FOLK$ (all rules) and $\FOMLL$ (rules in the dashed box)}
  \label{fig:LK1}
\end{figure}

@@

\todo{}

In this paragraph, we associate to each formula $A$ its \bfit{formula tree}
$\formtree{A}$, a directed
tree with leaves labelled by atoms, internal nodes labelled by connectives and
quantifiers, and edges directed from leaves to the root.

Let $\Gamma$ be a sequent in $\FOMLL$. Consider the formula tree
$\formtree{\form{\Gamma}}$. A \bfit{link} on $\Gamma$ is a pair of
leaves whose atoms are pre-dual. A \bfit{linking} $\lambda$ on $\Gamma$ is a set
of
disjoint links such that each leaf of $\Gamma$ is either labelled by $\ttt$ or
in exactly one link. Similar to the set of links in linked fographs, a linking can be seen as a
unification problem, and a dualizer of the linking $\lambda$ is an assignment
unifying all the links in $\lambda$. There exists a \bfit{most general dualizer}
 of $\lambda$ if $\lambda$ has a dualizer. \juihsuan{Now I use the same terminology as for linked fographs}
A \bfit{dependency} is a pair $(\single \exists x, \single \forall y)$ of nodes
such that the most general dualizer assigns to $x$ a term containing $y$.

Let $\lambda$ is a linking on $\Gamma$ that has a dualizer.
The \bfit{unification structure} $\unifstr{\lambda}$
associated to $\lambda$ is the formula tree $\formtree{\form{\Gamma}}$ together with an undirected
edge between leaves $l$ and $l'$ for every link $\{l, l'\}$ in $\lambda$ and a
directed edge from $\single \exists x$ to $\single \forall y$ for every
dependency $(\single \exists x, \single \forall y)$.

A \bfit{switching graph} of a unification structure $\unifstr{\lambda}$ is any
derivative of $\unifstr{\lambda}$ obtained by keeping only one edge into each $\cor$
and $\forall$ and undirecting remaining edges. A linking is \bfit{correct} if it
is unifiable and all of the switching graphs of its associated unification
structure are acyclic.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Translation between Unification Nets and $\FOMLL$}

\todo{}
\begin{thm}
  If a sequent is provable in $\FOMLL$, then there exists a unification net
  on it.
\end{thm}

\begin{proof}
  We proceed by induction on the proof of $\sqn \Gamma$ in $\FOMLL$, making a
  case analysis on the bottommost rule instance:
  \begin{itemize}
    \setlength\itemsep{.5em}
    \item $\vlinf{\axr}{}{\sqn{a,\cneg a}}{}$ : the linking $\{a, \dual{a}\}$ is correct.
    \item $\vlinf{\ttt}{}{\sqn{\ttt}}{}$ : the empty linking is correct.
    \item $\vliinf{\mixr}{}{\sqn{\Gamma,\Delta}}{\sqn\Gamma}{\sqn\Delta}$ :
	  By induction hypothesis, there is a correct linking on
	  $\Gamma$ and another one on $\Delta$, their union giving a
	  correct linking on $\Gamma, \Delta$.
    \item $\vlinf{\vlor}{}{\sqn{\Gamma,A\vlor B}}{\sqn{\Gamma,A,B}}$ : By
	  induction hypothesis, there is a correct linking on $\Gamma, A, B$,
	  and it is correct on $\Gamma, A \vlor B$ as well.
    \item $\vliinf{\vlan}{}{\sqn{\Gamma,A\vlan B, \Delta}}{\sqn{\Gamma,A}}{\sqn{B,\Delta}}$ :
          By induction hypothesis, there is a correct linking on $\Gamma, A$ and
	  another one on $B, \Delta$, their union giving a correct linking on
          $\Gamma,A \vlan B, \Delta$.
    \item $\vlinf{\exists}{}{\sqn{\Gamma,\exists x.A}}
	  {\sqn{\Gamma,A\ssubst{x}{t}}}$ : By induction hypothesis, there is a
	  correct linking $\lambda$ on $\Gamma, A\ssubst{x}{t}$.
	  For each atom in $\Gamma, A \ssubst{x}{t}$, there is a corresponding
	  atom in $\Gamma, \exists x.A$. There is therefore a linking $\lambda'$
          on $\Gamma, \exists x.A$ obtained from $\lambda$ via this
	  correspondence, and it is not difficult to check that $\lambda'$ is
	  correct as well.
    \item $\vlinf{\forall}{\text{($x$ not free in
	  $\Gamma$)}}{\sqn{\Gamma,\forall x.A}}{\sqn{\Gamma,A}}$ : By induction
          hypothesis, there is a correct linking on $\Gamma, A$, and it is easy
	  to see that it is a correct linking on $\Gamma, \forall x.A$ as well.

  \end{itemize}
\end{proof}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Translation between Unification Nets and Fonets}

@@

In $\FOMLL$, terms, atoms, formulas are defined as in first-order logic. For
simplicity, we choose to use $\cor$ and $\cand$ instead of $\parr$ and
$\otimes$ which are generally used in the presentation of linear logic.
A formula $A$ is identified with its \bfit{formula tree} $\formtree{A}$, a directed
tree with leaves labelled by atoms, internal nodes labelled by connectives and
quantifiers, and edges directed from leaves to the root. A \bfit{sequent} $\Gamma$ is simply a disjoint union of formulas. We write
comma for disjoint union.

@@

@@

\begin{definition} A \bfit{switching path} of a formula tree $\formtree{A}$ is a
	path in $\formtree{A}$ that does not go through both incoming edges of a $\cor$.
\end{definition}

@@

\begin{proposition}
\label{prop2}
Let $u$ and $v$ be two distinct vertices of a fograph $\graphof(A)$, then we have the equivalence
between:
\begin{itemize}
  \item $u$ and $v$ are adjacent in $\graphof(A)$
  \item $u$ and $v$ are connected by a switching path of $\formtree{A}$, and if one of them is a universal quantifier, then the other is
	not a descendant of the former.
  \end{itemize}
\begin{proof}
By induction on $A$.
\begin{itemize}
  \item If $A$ is an atom, trivial.
  \item If $A = A_1 \cand A_2$, then we distinguish two cases:
    \begin{itemize}
      \item $u$ and $v$ are both in $A_1$ (resp. $A_2$): trivial by
	      the induction hypothesis.
      \item one of them is in $A_1$ and the other is in $A_2$: they are
	      adjacent in $\graphof(A)$ by definition. By Proposition \ref{prop1},
		    the one in $A_1$ (resp. $A_2$) is connected to the
		    vertex representing $A_1$ (resp. $A_2$) by a switching
		    path. Together with the two edges incident to $A_1 \cand
		    A_2$, we obtain a switching path connecting $u$ and $v$.
    \end{itemize}
  \item If $A = A_1 \cor A_2$, then we distinguish two cases:
    \begin{itemize}
      \item $u$ and $v$ are both in $A_1$ (resp. $A_2$): trivial by
	      the induction hypothesis.
      \item one of them is in $A_1$ and the other is in $A_2$: they are not
	      adjacent in $\graphof(A)$ by definition. It is clear that they
		    are not connected by a switching path.
    \end{itemize}
  \item If $A = \exists x \ A'$, then we distinguish two cases:
    \begin{itemize}
      \item $u$ and $v$ are both in $A'$: trivial by the induction
	      hypothesis.
      \item one of them is $\exists x$ and the other is in $A'$: trivial
	      by Proposition \ref{prop1}
